---
title: "Using the poisson distribution to predict Premier League Matches"
output:
  pdf_document:
---

The aim of this program is to predict premier league matches using the poisson distribution. The Kelly Criterion will then be used in order to compare the predicted results with bet365's odds in order to see if there is any value in the bets. 

First a vector of all 20 EPL teams will be created, along with a value for the league average. 

```{r}
prem_teams<- c( "Arsenal", "Bournemouth", "Brighton", "Burnley", "Cardiff", "Chelsea", "Crystal Palace", "Everton", "Fulham", "Huddersfield", "Leicester", "Liverpool", "Manchester City", "Manchester United", "Newcastle","Southampton", "Tottenham", "Watford", "West Ham", "Wolves", "league average")
```

The read.csv function will then be used to import a csv file of each teams Average goals scored and conceded both home and away

```{r}
#Import data
prem_data = read.csv("prem_stats.csv", header = TRUE,
                     col.names = c("Team", "home_scored", "home_conceded",
                                   "away_scored", "away_conceded"), row.names = prem_teams)
#Remove first column
prem_data[,1]<- NULL

```

First lets check it's working by trying to retrieve some data
```{r}
prem_data
```

For each team ratings will be created for both their home and away attack ratings, as well as their home and away defence ratings. This will allow us to compare each team to the league average. The attack ratings will be created by dividing their average goals scored by the league average, similarly the defence ratings will be created by dividing their average goals conceded by the league average.
```{r}
#Adding attack and defence ratings to data frame
prem_data$home_attack_rating<- prem_data[,"home_scored"]/prem_data["league average","home_scored"]
prem_data$home_defence_rating<- prem_data[,"home_conceded"]/prem_data["league average", "home_conceded"]
prem_data$away_attack_rating<- prem_data[,"away_scored"]/prem_data["league average", "away_scored"]
prem_data$away_defence_rating<- prem_data[,"away_conceded"]/prem_data["league average", "away_conceded"]

```

Now we have our complete data frame let's have a look at some data
```{r}
#To view Arsenal's stats
prem_data["Arsenal",]

#To view all team's home attack ratings
prem_data[5]
```

Now we have our data frame working it is time to create our model
```{r}
#Define the function model
prem_model <- function(home_team, away_team, book_home,
                           book_away,book_draw, book_btts, book_over2.5)
  
{
  #Retrieves required information from the premiership stats data frame
  home_scored <- prem_data[home_team,"home_scored"]
  home_conceded <- prem_data[home_team, "home_conceded"]
  away_scored <- prem_data[away_team, "away_scored"]
  away_conceded <- prem_data[away_team, "away_conceded"]
  league_home_goals <- prem_data["league average", "home_scored"]
  league_away_goals <- prem_data["league average", "away_scored"]
  home_attack_rating <- prem_data[home_team, "home_attack_rating"]
  away_attack_rating <- prem_data[away_team, "away_attack_rating"]
  home_defence_rating <- prem_data[home_team, "home_defence_rating"]
  away_defence_rating <- prem_data[away_team, "away_defence_rating"]
  
  #Set the number of times the model will simulate each game
  num_trials = 1000000
  
  #create expected goals from data
  home_expected_goals <- home_attack_rating * away_defence_rating * league_home_goals
  away_expected_goals <- away_attack_rating * home_defence_rating * league_away_goals
  
  #Use poisson distribution to simulate how many goals each team will score
  home_goal_pred <- rpois(num_trials, home_expected_goals)
  away_goal_pred <- rpois(num_trials, away_expected_goals)
  
  #Calculate percentages of results
  total_goals <- home_goal_pred + away_goal_pred
  mean_goals <- mean(total_goals)
  home_win_percentage <<- (sum(home_goal_pred > away_goal_pred))/num_trials
  away_win_percentage <<- (sum(home_goal_pred < away_goal_pred))/num_trials
  draw_percentage <<- (sum(home_goal_pred == away_goal_pred))/num_trials
  over2.5_percentage <<- (sum(total_goals > 2))/num_trials
  btts_percentage <<- (sum(home_goal_pred > 0 & away_goal_pred > 0))/num_trials
  
  #Printing the expected outcomes
  cat("\nFor", home_team, "vs",away_team, "The expected outcomes are:", "\n","\n")
  
  cat(home_team, "Win Percentage:", round(home_win_percentage*100,2), "%","\n")
  cat(away_team, "Win Percentage:", round(away_win_percentage*100,2),"%")
  cat("\nDraw Percentage:", round(draw_percentage*100,2),"%")
  cat("\nOver 2.5 goals Percentage:", round(over2.5_percentage*100,2),"%")
  cat("\nBoth teams to score Percentage:", round(btts_percentage*100,2),"%", "\n")
  
  
  cat("\nThe implied odds are:","\n", "\n")
  
  #calculating the implied odds with the bookies edge included
  imp_home_odds<- round((1/home_win_percentage),2)
  imp_away_odds<- round((1/away_win_percentage),2)
  imp_draw_odds<- round((1/draw_percentage),2)
  imp_2.5_odds<- round((1/over2.5_percentage),2)
  imp_btts_odds<- round((1/btts_percentage),2)
  
  #printing these odds
  cat(home_team,":", imp_home_odds, "\n")
  cat(away_team,":", imp_away_odds)
  cat("\nDraw :", imp_draw_odds)
  cat("\nOver 2.5 goals :", imp_2.5_odds)
  cat("\nBoth teams to score :", imp_btts_odds, "\n")
  
  
  #Using Kelly Criterion to find the value of the bets
  home_wager_amount= round((((book_home - 1) * home_win_percentage) - (1 - home_win_percentage))/ (book_home-1),3)
  away_wager_amount= round((((book_away - 1) * away_win_percentage) - (1 - away_win_percentage))/ (book_away-1),3)
  draw_wager_amount= round((((book_draw - 1) * draw_percentage) - (1 - draw_percentage))/ (book_draw-1),3)
  over2.5_wager_amount= round((((book_over2.5 - 1) * over2.5_percentage) - (1 - over2.5_percentage))/ (book_over2.5-1),3)
  btts_wager_amount= round((((book_btts - 1) * btts_percentage) - (1 - btts_percentage))/ (book_btts-1),3)
  
  cat("\nThe Value of the bets are:")
  
  cat("\nThe Value in betting on",home_team,"is:", home_wager_amount*100, "%")
  cat("\nThe Value in betting on",away_team,"is:", away_wager_amount*100, "%")
  cat("\nThe Value in betting on the draw is:", draw_wager_amount*100, "%")
  cat("\nThe Value in betting on over 2.5 goals is:", over2.5_wager_amount*100, "%")
  cat("\nThe Value in betting on both teams to score is", btts_wager_amount*100, "%")
  
  
  
}
```

Now the function has been imported it can be used on the premier league fixtures to find expected results and the implied oddss from this, as well as the value of the bet. For example for Chelsea vs Man United:
```{r}
#Chelsea vs Manchester United
prem_model("Chelsea", "Manchester United",
          book_home = 1.75, book_away = 5.25, book_draw = 3.85,
          book_btts = 1.75, book_over2.5 = 1.80)
```

Here we can see our model gives the odds for over 2.5 goals to be 1.25 (as 79% of our one million simulations ended with over 2.5 goals) whilst bet365 are offering odds of 1.90. This is then converted using the Kelly Criterion to suggest we stake 54.7% of our bankroll on over 2.5 goals.


The model can be used to run all fixtures in a weekend.
```{r}
#Arsenal vs Leicester
prem_model("Arsenal", "Leicester",
          book_home = 1.53, book_away = 6.50, book_draw=3.60,
          book_btts = 1.57, book_over2.5 = 1.44)

#Cardiff vs Fulham
prem_model("Cardiff", "Fulham",
           book_home = 2.55, book_away = 2.90, book_draw = 3.50,
           book_btts = 1.66, book_over2.5 = 1.80)

#Manchester City vs Burnley
prem_model("Manchester City", "Burnley",
           book_home = 1.083, book_away = 34.00, book_draw = 13.00,
           book_btts = 2.50, book_over2.5 = 1.28)

#Newcastle vs Brighton
prem_model("Newcastle", "Brighton",
           book_home = 2.20, book_away = 3.80, book_draw = 3.25,
           book_btts = 2.05, book_over2.5 = 2.50)

#West Ham vs Tottenham
prem_model("West Ham", "Tottenham",
           book_home = 4.10, book_away = 1.90, book_draw = 3.90,
           book_btts = 1.57, book_over2.5 = 1.66)

#Wolves vs Watford
prem_model("Wolves", "Watford",
           book_home = 1.75, book_away = 5.25, book_draw = 3.75,
           book_btts = 1.95, book_over2.5 = 2.00)

#Huddersfield vs Liverpool
prem_model("Huddersfield", "Liverpool",
           book_home = 11.00, book_away = 1.33, book_draw = 5.50,
           book_btts = 2.25, book_over2.5 = 1.80)

#Everton vs Crystal Palace
prem_model("Everton", "Crystal Palace",
           book_home = 1.85, book_away = 4.75, book_draw = 3.60,
           book_btts = 1.80, book_over2.5 = 2.00)

#Bournemouth vs Southampton
prem_model("Bournemouth", "Southampton",
           book_home = 2.05, book_away = 3.75, book_draw = 3.70,
           book_btts = 1.66, book_over2.5 = 1.80)
```

We can see that the best value bets are: Chelsea to beat Man United and over 2.5 goals, Crystal Palace to beat Everton, Liverpool to beat Huddersfield and Tottenham to beat West Ham. 





